Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-24 manuscript.rmd
1. Edited abstract 2. Moved methods summary text to ME&M 3. Edited intro to ERD model

- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching critical safety thresholds. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD, as the average depth of water extraction), for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and to leaf hydraulic vulnerability curves. We validated ERD against existing tree and soil stable water isotope data. 
- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts. Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
- The role of deep-water access in alleviating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.

@@

\
\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

\
\     We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig. \@ref(fig:schematic)). This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD, given the corresponding gradient in isotopic composition of water in the soil profile. We tested relationship of ERD with above-ground hydraulic traits,namely, xylem vulnerability to embolism, branch hydraulic safety margins, turgor loss points and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets. Finally, ee analysed species' mortality rates across the censuses spanning 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

\
Fig. \@ref(fig:schematic) here.
Table \@ref(tab:hypo) here.

# Materials and Methods

@@

To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes. In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) [@Bonan:2013]. FATES is a community-based, open-source model used for studying climate-vegetation interactions. FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions for ELM and FATES can be found elsewhere [@Fisher.2010; @Fisher.2015; @Koven:2020].

@@

\
\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 (Faybishenko et al. 2020) initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

@@

To estimate effective rooting depth, we tested alternate models for growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. In this study, we used maximum diurnal leaf conductance ($K_\textrm{leaf}$) approximated using rhizosphere soil water potential ($\Psi$~soil~) as a proxy for the soil water limiting factor on growth. In our models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. For each species', growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\
\    As VPD may have a non-linear relationship with growth, with growth increasing with VPD intially, but decreasing with VPD beyond an optimal value due to stomatal closure [AT:Chonggang, add a reference], we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

@@

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\Psi_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\Psi_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood (or, the minimum AIC score), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $R$^2 >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

@@

\
\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought. The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa). The difference in the results shown here are likely attributable to site differences in the precipitation regime. In the dry forest site in CT2018 (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The kind of droughts that buffered the deep rooted species in our study were recurrent sub-annual droughts that extended the dry season into the wet season. The variation in mortality explained by ERD in our study is upto 40%, and other causes may explain observed mortality [@McDowell.2018].

@@

The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests, which intrinsically rely on the parameterization of contrasting life history strategies [@Schieter:2012] and the simulation of competition between those strategies. The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off ($n$ = 26 species) at our study site stem level. We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). This offers the promise of a greater ability to parameterie the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics. The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

>>>>

- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins can predict mortality risk at given levels of dehydration, but deep-water access may delay plants from dehydration. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD, as the average depth of water extraction), for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and to leaf hydraulic vulnerability curves. We validated ERD against existing tree and soil stable water isotope data. 
- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates during extreme droughts. Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
- The role of deep-water access in alleviating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.

@@

\
\     Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in terms of hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region.

\
\   We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

Table \@ref(tab:hypo) here.

# Materials and Methods
\
\      We developed a novel, empirical inverse model to estimate species effective rooting depths (ERD) for large trees of `r length(erd.sp)` species (Fig. \@ref(fig:schematic)). We defined species ERD as the depth at which soil water potential best explained species-specific 5-yrly mean growth rates over 25 years. To obtain the daily dynamics of soil water potential in the whole-soil column we locally parameterized and calibrated a 1-D hydrological water balance model. The ERD model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by using existing data for eight species and developing trait-based proxies for the rest. We evaluated ERD against existing dueterium to hydrogen content data for a subset of six tree species as an independent observation of ERD. We tested relationship of ERD with above-ground hydraulic traits, namely, xylem vulnerability to embolism, branch hydraulic safety margins, leaf turgor loss point and maximum stem-specific hydraulic conductivity, for a subset of eight species with both datasets. Finally, we analysed species' mortality rates over 35 years to test whether ERD explained risk of drought-induced mortality in periods of extreme hydrological drought.

\
Fig. \@ref(fig:schematic) here.

@@

To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes. In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) similar to that in CLM4.5 [@Bonan:2013]. FATES is a community-based, open-source model used for studying climate-vegetation interactions. FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions for ELM and FATES can be found elsewhere [@Fisher.2010; @Fisher.2015; @Koven:2020].

@@

\
\   We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers measured at the BCI meteorological station for the period from 1985-2018 (Faybishenko et al. 2020) initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was on deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

@@

To estimate effective rooting depth, we tested alternate models for growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen.2018]. In moist forests $\Psi$~leaf~ equilibriates with $\Psi$~soil~ at night and declines through the day as plants transpire. $\Psi$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012]. We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\Psi$~soil~ as a proxy for the soil water limitation on growth. In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. This approach ignores the buffering effect of stem water storage on $K_\textrm{leaf}$, which may be a subject of future work (see Discussion). VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020]. The fraction of a tree's leaf cover present at a given time further limits it's growth potential. We therefore regressed species-specific growth in a census interval against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\
\    To account for non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

@@

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\Psi_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\Psi_{\textrm{soil,} z}$) best explained observed dynamics of species growth (six 5-yrly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood (or, the minimum AIC score), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $R$^2 >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

@@

\
\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought. The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa). The difference in the results shown here are likely attributable to site differences in the precipitation regime. In the dry forest site in CT2018 (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The kind of droughts that buffered the deep rooted species in our study were recurrent sub-annual droughts that extended the dry season into the wet season. The variation in mortality explained by ERD in our study is upto 40%, and other causes may explain observed mortality [@McDowell.2018].

@@

The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests, which intrinsically rely on the parameterization of contrasting life history strategies [@Schieter:2012] and the simulation of competition between those strategies. We found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). This offers the promise of a greater ability to parameterie the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics. The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.
